---
title: "Install R packages"
output: 
  rmarkdown::html_document:
    fig_width: 6
    fig_height: 4
    theme: flatly
    mathjax: NULL
    highlight: pygments
    self_contained: false
    pandoc_args: ["--mathjax"]
editor_options: 
  chunk_output_type: console
---


<style>
pre.sourceCode {
  background-color: #fcffff;
  color: #262c2d;
}
code span.at {
  color: #ca863c;
}
code span.dv {
  color: #047d39;
}
code span.fu {
  color: #7629d2;
}
code span.do {
  color: #999;
  font-style: normal;
}
code span.st {
  color: #cc4e11;
}
code span.sc {
  color: #ca7dac;
}
pre > code.sourceCode a {
  display: none;
}
blockquote > p {
  font-size: 13px !important;
  color: #647980;
  line-height: 1.5 !important;
  font-weight: 400 !important;
}
blockquote > code {
   rgba(0, 0, 0, 0.02);
}
</style>


```{r, include = FALSE}
ECICourse::upload_images_to_canvas()
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  collapse = TRUE
)
```


Perform the following steps to ensure you have the necessary R packages for this course installed. 

## Step 1

Open RStudio

## Step 2

Ensure these packages are installed, and install them if not.  Copy and paste the code below into the R console:
```{r message = FALSE, warning = FALSE}
if (!require("tidyverse")) {
  install.packages("tidyverse")
  library("tidyverse")
}
if (!require("coin")) {
  install.packages("coin")
  library("coin")
}
if (!require("DeclareDesign")) {
  install.packages("DeclareDesign")
  library("DeclareDesign")
}
if (!require("dagitty")) {
  install.packages("dagitty")
  library("dagitty")
}
if (!require("Matching")) {
  install.packages("Matching")
  library("Matching")
}
if (!require("AER")) {
  install.packages("AER")
  library("AER")
}
```


## Step 3

Install the ECICourse package. Copy and paste the code below into the R console:

```{r message = FALSE, warning = FALSE}
if (!require("remotes")) {
    install.packages("remotes")
}
if (!require("ECICourse")) {
    remotes::install_github("jasonmtroos/ECICourse")
    library("ECICourse")
}
```

Note that this package changes every year. If you are repeating this course and previously installed the course package, you must re-install it now.

## Later in the course

I might need to update the ECICourse package during the course. If that happens, I will create an announcement asking you to re-install the package. To re-install the package, you will need to run the following code:

```{r message = FALSE, warning = FALSE, eval = FALSE}
unloadNamespace('ECICourse')
remotes::install_github("jasonmtroos/ECICourse")
```

