---
title: "Install R packages"
output: 
  rmarkdown::html_document:
    fig_width: 6
    fig_height: 4
    mathjax: null
    theme: cerulean
    highlight: pygments
    self_contained: false
editor_options: 
  chunk_output_type: console
---

<style>
pre > code.sourceCode a {
  display: none;
}
pre {
  border-width: 0px;
}
code {
  border-width: 0px;
}
</style>

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


Perform the following steps to ensure you have the necessary R packages for this course installed. When you are done, click on "Mark as done" in the top-right corner of this page.

## Step 1

Open RStudio

## Step 2

Ensure these packages are installed, and install them if not.  Copy and paste the code below into the R console:
```{r message = FALSE, warning = FALSE}
if (!require("tidyverse")) {
  install.packages("tidyverse")
  library("tidyverse")
}
if (!require("coin")) {
  install.packages("coin")
  library("coin")
}
if (!require("DeclareDesign")) {
  install.packages("DeclareDesign")
  library("DeclareDesign")
}
if (!require("dagitty")) {
  install.packages("dagitty")
  library("dagitty")
}
if (!require("Matching")) {
  install.packages("Matching")
  library("Matching")
}
if (!require("AER")) {
  install.packages("AER")
  library("AER")
}
```


## Step 3

Install the ECICourse package. Copy and paste  the code below into the R  console:

```{r message = FALSE, warning = FALSE}
if (!require("remotes")) {
    install.packages("remotes")
}
if (!require("ECICourse")) {
    remotes::install_github("jasonmtroos/ECICourse")
    library("ECICourse")
}
```


## Later in the course

I will very likely need to update the ECICourse package during the course. Whenever that happens, I will create an announcement asking you to reinstall the package. To reinstall the package, you will need to run the following code:

```{r message = FALSE, warning = FALSE, eval = FALSE}
unloadNamespace('ECICourse')
remotes::install_github("jasonmtroos/ECICourse")
```

