---
title: "R Tutorial: Calculating average treatment effects"
output: 
  rmarkdown::html_document:
    fig_width: 6
    fig_height: 4
    mathjax: null
    theme: cerulean
    highlight: pygments
    self_contained: false
editor_options: 
  chunk_output_type: console
---


<style>
pre > code.sourceCode a {
  display: none;
}
pre {
  border-width: 0px;
}
code {
  border-width: 0px;
}
</style>


```{r, include = FALSE}
if (require(canvasapicore)) {
  knitr::opts_knit$set(upload.fun = function(file) {
    canvasapicore::load_token_and_domain()
    resp <- cnvs::cnvs_upload("/api/v1/folders/607253/files", file)
    structure(resp$url, XML = resp)
  })
}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


Read along with this tutorial, and run the R code in RStudio as you go (copy and paste, or type directly into R). After you have completed this tutorial, mark this task as done by clicking on "Mark as done" in the top-right corner of this page.

Start by loading packages:

```{r setup}
library(ECICourse)
library(tidyverse)
```

The  data for this tutorial describe a fictional field  experiment  conducted by an educational software company in cooperation with a local university. Students in the same degree program were randomized into two  conditions before entering  a required, core course. In the  treatment condition, students received access to the educational software. In the control condition, students did not  gain access to the software.

In this tutorial, you will go  through the steps needed to calculate the  average treatment  effect (ATE) of having access  to the educational software. 

## Obtain the data

To obtain the data describing this experiment, run the following code:

```{r}
fieldexp <- get_ECI_data('module-2-tutorial')
```

The variable `fieldexp` is a data frame with `r ncol(fieldexp)` columns and `r nrow(fieldexp)` rows. Each row describes a  student. The column  `D` indicates whether the student was assigned to the treatment condition (`D == 1`) or to the control condition (`D == 0`). The column `Y` reflects the student's grade at the end of the course.

Look at the first few observations:

```{r}
fieldexp
```

Half the students were assigned to the treatment and half to the control condition:

```{r}
fieldexp %>%
  count(D)
```

Look at the distribution of grades in the  two conditions:

```{r M2-dist-DY}
ggplot(fieldexp, aes(x = factor(D), y = Y)) +
  geom_jitter(width = .1, height = .1) +
  labs(x = 'D')
```

## Average grades in the treatment and control conditions

What were the average  grades among students who had access to the software (the treatment group,  or those with `D == 1`). Conceptually, this is just the average value of `Y` among students who had `D == 1`, or the mean  of `Y` conditional on `D == 1`. 


To calculate the average outcome (grade) among students who received access to the software (students with `D == 1`), we first filter the data frame, keeping only rows with `D == 1`, and then we calculate `mean(Y)`:

```{r}
fieldexp %>%
  filter(D == 1) %>%
  summarise(Y = mean(Y))
```

To calculate  the average   grade among students who did not receive the  software (the  control group, those with `D == 0`),  we follow a similar procedure:

```{r}
fieldexp %>%
  filter(D == 0) %>%
  summarise(Y = mean(Y))
```

We can use `group_by` to calculate both of these averages at the same time. In the code below, we  summarize the mean of `Y` separately for each of the two groups defined by the two values of `D`.  We will also  use `mutate` to assign descriptive labels to the values in `D`  (renaming `D == 0` to `'Control'` and `D == 1` to `'Treatment'`).

```{r}
fieldexp %>%
  group_by(D) %>%
  summarise(Y = mean(Y)) %>%
  mutate(D = case_when(D == 0 ~ 'Control',
                       D == 1 ~ 'Treatment'))
```

Finally, we can use  `pivot_wider` to transform the  output, creating two new columns based on the contents of each row:

```{r}
fieldexp %>%
  group_by(D) %>%
  summarise(Y = mean(Y)) %>%
  mutate(D = case_when(D == 0 ~ 'Control',
                       D == 1 ~ 'Treatment')) %>%
  pivot_wider(names_from = 'D', values_from = 'Y')
```

## Average  treatment effect

The average treatment effect (ATE) is ð”¼[Y(1) -  Y(0)], or equivalently, ð”¼[Y(1)] - ð”¼[Y(0)]. We don't observe the potential outcomes Y(1) and Y(0), so we assume that the conditional averages of `Y` when  `D == 1`  and `D == 0` (calculated above) are reasonable  estimators for ð”¼[Y(1)] and ð”¼[Y(0)]. This assumption allows us to estimate of the average treatment effect by simply taking the difference between the average values of `Y` when `D` is either  1 or 0. 

```{r}
fieldexp %>%
  group_by(D) %>%
  summarise(Y = mean(Y)) %>%
  mutate(D = case_when(D == 0 ~ 'Control',
                       D == 1 ~ 'Treatment')) %>%
  pivot_wider(names_from = 'D', values_from = 'Y') %>%
  mutate(ATE = Treatment - Control)
```

The average treatment effect is the value in the column labeled `ATE`. Therefore, the ATE for  this experiment is `r round(lm(Y ~ D,  data = fieldexp)$coef[2], 3)`. This is the expected  difference  in students' grades caused by having access to the educational software.


