---
title: "R Tutorial: Classical matching"
output: 
  rmarkdown::html_document:
    fig_width: 6
    fig_height: 4
    mathjax: null
    theme: cerulean
    highlight: pygments
    self_contained: false
editor_options: 
  chunk_output_type: console
---

<style>
pre > code.sourceCode a {
  display: none;
}
pre {
  border-width: 0px;
}
code {
  border-width: 0px;
}
</style>


```{r, include = FALSE}
# if (require(canvasapicore)) {
#   knitr::opts_knit$set(upload.fun = function(file) {
#     canvasapicore::load_token_and_domain()
#     resp <- cnvs::cnvs_upload("/api/v1/folders/601289/files", file)
#     structure(resp$url, XML = resp)
#   })
# }
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


Read along with this tutorial, and run the R code in RStudio as you go (copy and paste, or type directly into R). After you have completed this tutorial, mark this task as done by clicking on "Mark as done" in the bottom-right corner of this page.

Start by loading packages:

```{r setup, cache = FALSE}
library(ECICourse)
library(tidyverse)
```

The data for this tutorial describe a fictional observational study  conducted by an online seller of accessories for smart phones and tablet computers. The company emailed a single-use coupon code to all `N = 20000` customers in its database. The coupons, which were valid on orders of €5 or more, could be used to receive a 30% discount on the customers' next order, as long as they were redeemed within the next month.

The company wants to learn how much order sizes increased as a result of customers using the coupons. However, because the customers chose whether to use the coupon or not, there is selection. The company therefore realizes that the naive estimator of the average treatment effect will be biased and inconsistent. 

In this tutorial, we will use a classical matching strategy to calculate the average treatment effects among the treated (ATT; those using the coupons on their orders), controls (ATC; those who did not use the coupon), and the average customer (ATE).


## Data

To obtain the data describing this experiment, run the following code:

```{r}
coupon <- get_ECI_data('module-6-tutorial')
```

The data contain an id for each customer (`customer`), background information describing each  customer (see below), an indicator of whether the customer redeemed the coupon (`used_coupon`), and the size of the first order the customer placed during the promotion (not including the discount, `order_size`). If no order was placed, then `order_size == 0`. To simplify this tutorial, we will assume that none of the customers placed more than one order  during the month of the promotion.

```{r}
coupon
```

The columns `recency`, `frequency`, and `monetary`  are  binary variables describing the past purchase behavior of each customer:

- The value of `recency` is set to `Overdue` if, according the company's predictive sales model, the customer is likely to place an order soon. Otherwise, the value of `recency` is set to `Recent`, meaning the customer placed an order recently enough that the company does not expect another anytime soon.

- The value of `frequency` is set to `Frequent` if the customer places orders with relatively high frequency. Otherwise, the value of `frequency` is set to `Infrequent`.

- The value of `monetary` is set to `High value` if the customer's average order size and margin are relatively high. Otherwise, the value of `monetary` is set to `Low value.`

The rate of coupon use varied across customers with different values of `recency`, `frequency`, and `monetary`. 

```{r}
coupon %>%
  group_by(recency, frequency, monetary) %>%
  summarise(used_coupon = mean(used_coupon)) 
```

The output above shows that customers who placed an order recently, tend to order less frequently,  and have low monetary value redeemed the coupon at a rate of about 1%. Customers who were overdue for an order, order more frequently, and place high-value orders redeemed the coupon at a much higher rate: about 60%.

Finally, notice that some customers placed orders *without* using their coupon. These customers might have deleted the email without reading it.

```{r}
coupon %>%
  group_by(recency, frequency, monetary, used_coupon) %>%
  summarise(n_who_placed_order = sum(order_size > 0))
```

## Conditional average treatment effects

The company believes that the stratifying variables `recency`, `frequency`, and `monetary` completely account for all systematic variation in whether customers used the coupon. In other words, the company is willing to assume that all selection is on the *observables*. This means that among groups of customers in the same stratum---i.e., customers who have all three values of `recency`, `frequency`, and `monetary` in common---we are assuming that coupon use was essentially random. This is a *very* strong assumption to make, but it is what we need to assume in order to use a classical matching procedure to estimate average treatment effects.

To help build intuition for how matching works, we will first consider just a single stratum of customers with `recency == 'Overdue'`, `frequency == 'Frequent'`, and `monetary == 'High value'`:

```{r}
coupon %>%
  filter(recency == 'Overdue' & frequency == 'Frequent' & monetary == 'High value')
```

```{r echo = FALSE, results = 'hide'}
high_S <- coupon %>%
  filter(recency == 'Overdue' & frequency == 'Frequent' & monetary == 'High value')
```

The assumption of selection on the observables means that within this group, coupon use is independent of *potential* outcomes---how large orders *would be* when using or not using a coupon. Under this assumption, any difference in average order size among coupon users and non-users (in this stratum) is due to the coupon.

```{r}
coupon %>%
  filter(recency == 'Overdue' & frequency == 'Frequent' & monetary == 'High value') %>%
  group_by(used_coupon)  %>%
  summarise(avg_order_size = mean(order_size), n_customers = n())
```

The conditional average treatment effect (CATE) for this stratum of customers is difference between these two average order sizes.

```{r}
coupon %>%
  filter(recency == 'Overdue' & frequency == 'Frequent' & monetary == 'High value') %>%
  group_by(used_coupon)  %>%
  summarise(avg_order_size = mean(order_size)) %>%
  pivot_wider(names_from = 'used_coupon', values_from =  'avg_order_size',
              names_prefix = 'coupon_') %>%
  mutate(CATE = coupon_1 - coupon_0) 
```

Following this logic, we can calculate CATEs for *all* strata---i.e., for all unique combinations of the three stratifying variables (`recency`, `frequency`, and  `monetary`). First, we calculate the average order size among coupon users and non-users in each stratum:

```{r}
avg_order_sizes_by_group <- 
  coupon %>%
  group_by(recency, frequency, monetary, used_coupon) %>%
  summarise(avg_order_size = mean(order_size)) %>%
  ungroup()
avg_order_sizes_by_group
```

We can now calculate the CATE for each group:

```{r}
avg_order_sizes_by_group %>%
  pivot_wider(names_from  = 'used_coupon', values_from = 'avg_order_size', 
              names_prefix = 'coupon_') %>%
  mutate(CATE = coupon_1 - coupon_0) %>%
  rename(with_coupon = coupon_1, without_coupon = coupon_0) # renaming columns to 
                                                            # make it easier to read
```

The estimated CATEs shown above vary widely across groups of customers. As we will see below, one way to think of classical matching is that it works by taking weighted averages of these CATEs in order to generate unbiased and consistent estimates for the average treatment effect (ATE), average treatment effect on the treated (ATT), and the average treatment effect on the controls (ATC). The weights in the weighted averages depend on the target effect to be estimated. 

## Estimating the ATT with classical matching

First, we will use a matching procedure to estimate the ATT. For these data, the ATT is the average effect of coupon use on customers who chose to use the coupon. If that's confusing, think instead of the ATT as an answer to the following question: "Among customers who chose to use the coupon, how much smaller would their orders have been if they hadn't used the coupon?"

In the material below, we will first apply the matching procedure in a way that goes slowly and uses a lot of code, but is conceptually more instructive. At the end I will show you an equivalent procedure, using the Matching R package, that requires less code.

### Step by step

First, isolate the treatment and control groups into different data frames:

```{r}
treatment <-
  coupon %>%
  filter(used_coupon == 1)
control <-
  coupon %>%
  filter(used_coupon == 0)
```

The two data frames are not balanced, by which we mean that within each stratum of the variables  `recency`, `frequency`, and `monetary`, there are a different number of coupon users and non-users.

```{r}
treatment %>%
  count(recency, frequency, monetary)
control %>%
  count(recency, frequency, monetary)
```

To calculate the ATT, we want to construct an *artificial* control group that is balanced with respect to the treatment group. Consider this example: There are `r sum(high_S$used_coupon)` treated customers with `recency == 'Overdue'`, `recency == 'Frequent'`, and `monetary == 'High value'`. 


```{r}
n_treated_in_this_stratum <- 
  treatment %>%
  filter(recency == 'Overdue' & frequency == 'Frequent' & monetary == 'High value') %>%
  nrow()
n_treated_in_this_stratum
```

To calculate  the ATT, the artificial control group for this stratum needs to contain exactly `r sum(high_S$used_coupon)` coupon non-users. But because the *actual* number of customers in the control group (for this stratum) is less than `r  n_treated_in_this_stratum` (recall there are `r nrow(high_S) - n_treated_in_this_stratum` customers in this stratum who did not use the coupon), we need to sample these customers *with replacement* from the original control group when constructing the artificial control. After sampling, the artificial control group for this stratum might look something like this:

```{r echo = FALSE, results = 'hide'}
set.seed(1)
```

```{r}
artificial_control_for_this_stratum <- 
  control %>%
  filter(recency == 'Overdue' & frequency == 'Frequent' & monetary == 'High value') %>%
  sample_n(n_treated_in_this_stratum, replace = TRUE) %>%
  arrange(customer)
artificial_control_for_this_stratum
```

Because we sampled with replacement from the actual control group, many customers are repeated in this artificial control group. 

To calculate the ATT for the entire customer base, we need to repeat this sampling procedure for each stratum (i.e., unique combination of values for the three variables). To do that, we need to know how many *treated* customers there are in each stratum. We will calculate that first:

```{r}
n_treated_per_stratum <-
  treatment %>%
  count(recency, frequency, monetary, name = 'n_treated')
n_treated_per_stratum
```

Next, within each stratum, we need to sample `n_treated` customers from the control group. The approach here will be to join the tables `control` (containing control group customers) and `n_treated_per_stratum` (indicating how many control customers to sample from each stratum). We will  then use the values of `n_treated` and the `sample_n` function (from dplyr) to construct the artificial control group.

```{r}
control_with_n_treated <- 
    inner_join(control, n_treated_per_stratum, 
               by = c("recency", "frequency", "monetary"))
control_with_n_treated

artificial_control <-
  control_with_n_treated %>%
  group_by(recency, frequency, monetary, n_treated) %>%
  sample_n(size = n_treated[1], replace = TRUE) %>%       #  within each stratum, sample n_treated
  ungroup() %>%                                           #    controls with replacement
  select(-n_treated)
```

The `treatment` and `artificial_control` data frames are now balanced within each stratum:

```{r}
treatment %>%
  count(recency, frequency, monetary)
artificial_control %>%
  count(recency, frequency, monetary)
```

Now that we have an artificial control group that is balanced with respect to the treatment group, the next step is to combine these so we can calculate the conditional (within stratum) average treatment effect on the treated (CATT):

```{r}
treatment_and_artificial_control <-
  bind_rows(treatment, artificial_control)
CATTs <- 
  treatment_and_artificial_control %>%
  group_by(recency, frequency, monetary, used_coupon) %>%
  summarise(avg_order_size = mean(order_size))  %>%
  ungroup() %>%
  pivot_wider(names_from = 'used_coupon', values_from = 'avg_order_size',
              names_prefix = 'coupon_') %>%
  mutate(CATT = coupon_1 - coupon_0) %>%
  select(-coupon_0, -coupon_1)
CATTs
```

The final step in calculating the ATT is to take a weighted average of the values of `CATT`. The weight for each stratum is proportional to the number of *treated* customers in that stratum:

```{r}
CATT_with_n_treated_per_stratum <- 
  CATTs %>%
  inner_join(n_treated_per_stratum, 
             by = c("recency", "frequency", "monetary"))
CATT_with_n_treated_per_stratum
CATT_with_n_treated_per_stratum  %>%
  summarise(ATT = sum(n_treated * CATT) / sum(n_treated))
```

### A more concise approach using dplyr

The  code above is spread out with a lot of discussion. Here is a version that is more compact, and makes use of the `group_map` function in dplyr. Feel free to skip over this material and jump to the next subsection (showing how to use the Matching package).

```{r}
coupon %>%
  group_by(recency, frequency, monetary) %>%
  mutate(n_treated = sum(used_coupon)) %>%   # number of customers in the treated condition within each stratum
  group_by(recency, frequency, monetary, 
           used_coupon, n_treated) %>%
  group_map(                                 # apply the function below to each group_by subset of rows
    ~sample_n(.x,                            # sample rows from this group_by subset (.x)
              size = .y$n_treated,           # the number of rows to sample = n_treated (.y refers to the grouping variables)
              replace = !.y$used_coupon),    # sample with replacement if we are sampling from controls, otherwise
    .keep = TRUE) %>%                        #   sample *without* replacement if we are sampling from treatment so we
  bind_rows() %>%                            #   get back the original treated subset
  group_by(recency, frequency, monetary,     # calculate average order size within stratum and treatment/control
           used_coupon, n_treated) %>%
  summarise(
    avg_order_size = mean(order_size)) %>%
  ungroup() %>%
  pivot_wider(names_from = 'used_coupon',    # finally, calculate ATT as a weighted average of CATEs
              values_from = 'avg_order_size',
              names_prefix = 'coupon_') %>%  
  summarise(ATT = sum((coupon_1 - coupon_0) * n_treated)/sum(n_treated))
```

### An even *more* concise approach using the Matching package

First, load the Matching package. The Matching package also loads the MASS package, which replaces dplyr's version of `select`. We will need dplyr's `select`, however, so we call `select <- dplyr::select` after loading packages. 

```{r}
library(Matching)
select <- dplyr::select     # because MASS replaced dplyr::select and we want the dplyr version
```

To calculate the ATT using the Matching package, we need to create distinct variables for 1) the outcome variable (`order_size`), 2) the matching variables (`recency`, `frequency`, and `monetary`), and 3) the treatment variable (`used_coupon`):

```{r}
Y <- coupon$order_size
X <- 
	coupon %>%
	select(recency, frequency, monetary) %>%
	mutate(across(everything(), unclass)) %>%     # Matching expects a matrix of numbers, not a data frame with factors
	as.matrix()
D <- coupon$used_coupon
att <- Match(Y = Y, Tr = D, X = X, 
					 estimand = 'ATT', M = 1, 
					 replace = TRUE, ties = FALSE)
summary(att)
```

Notice that the `Match` procedure calculates a standard error for the estimate, which we didn't do in our version based on dplyr. 

### ATC and ATE using dplyr

To calculate the ATC, we follow a very similar procedure, except we sample *treatment* cases in order to create an *artificial treatment group* that matches the number of control units in each stratum. Hence, code below is similar to what we did above. The main differences are that we now we count `n_control = sum(used_coupon == 0)`, and we sample with replacement from the treatment group (when `used_coupon == 1`):

```{r}
control_and_artificial_treatment <- 
  coupon %>%
  group_by(recency, frequency, monetary) %>%
  mutate(n_control = sum(used_coupon == 0)) %>%
  group_by(recency, frequency, monetary, used_coupon, n_control) %>%
  group_map(
    ~sample_n(.x, size = .y$n_control, 
              replace = .y$used_coupon), .keep = TRUE) %>%
  bind_rows()
```

Because most customers did not use the coupon, the resulting data frame has more observations than the original (the number of observations is exactly twice the number of customers who did not use the coupon):

```{r}
nrow(control_and_artificial_treatment)
nrow(coupon)
```

To calculate the ATC we continue in a similar manner as before:

```{r}
control_and_artificial_treatment %>%
  group_by(recency, frequency, monetary, used_coupon, n_control) %>%
  summarise(avg_order_size = mean(order_size)) %>%
  ungroup() %>%
  pivot_wider(names_from = 'used_coupon', values_from = 'avg_order_size',
              names_prefix = 'coupon_') %>%
  summarise(ATC = sum((coupon_1 - coupon_0) * n_control)/sum(n_control))
```

The ATC for  this study represents  the increase in order size we would expect among customers who didn't use the coupon, if we could somehow force them to place an order using the coupon. 

To calculate the ATE, we can take a weighted average of the estimated ATT and ATC, with weights determined by the share of coupon users and non-users. Something like: `(13.48 * 3236 + 9.77 * 16764) / 20000` which  is about `r round((13.48 * 3236 + 9.77 * 16764) / 20000, 2)`.

### ATC and ATE using the Matching package

As an alternative, we can simply use the Matching package to calculate the ATC and ATE, just as we did when calculating the ATT.

```{r}
atc <- Match(Y = Y, Tr = D, X = X, 
					 estimand = 'ATC', M = 1, 
					 replace = TRUE, ties = FALSE)
summary(atc)
ate <- Match(Y = Y, Tr = D, X = X, 
					 estimand = 'ATE', M = 1, 
					 replace = TRUE, ties = FALSE)
summary(ate)
```

