---
title: "R Tutorial: Estimating the CACE/LATE"
output: 
  rmarkdown::html_document:
    fig_width: 6
    fig_height: 4
    mathjax: null
    theme: cerulean
    highlight: pygments
    self_contained: false
editor_options: 
  chunk_output_type: console
---

<style>
pre > code.sourceCode a {
  display: none;
}
pre {
  border-width: 0px;
}
code {
  border-width: 0px;
}
</style>


```{r, include = FALSE}
if (require(canvasapicore)) {
  knitr::opts_knit$set(upload.fun = function(file) {
    canvasapicore::load_token_and_domain()
  	file <- stringr::str_replace(file, "Dropbox (Erasmus Universiteit Rotterdam)", "Dropbox_RSM")
    resp <- cnvs::cnvs_upload(file, "/api/v1/folders/1191228/files")
    structure(resp$url, XML = resp)
  })
}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


Read along with this tutorial, and run the R code in RStudio as you go (copy and paste, or type directly into R). After you have completed this tutorial, mark this task as done by clicking on "Mark as done" in the bottom-right corner of this page.

Start by loading packages:

```{r setup, cache = FALSE}
library(ECICourse)
library(tidyverse)
```

## Data

The data for this tutorial describe the outcome of a fictional experiment involving an incentive to join a customer loyalty program. A software company produces mobile apps using a "freemium" sales model: the base app is available for free, and add-ons sell for €.50 each. The company announced a customer loyalty program. Customers would be able to enroll in the loyalty program for free, and in return, they would gain access to early releases of new add-ons, as well as discounts on (as of yet unspecified) future products. Because the company sells its app via mobile app stores, it knows the following:

* Who each customer is
* Whether they have enrolled in the loyalty program
* How much revenue each customer generates

The company decided to test incentives for customers to join the loyalty program. Just prior to the launch of the loyalty program (i.e., before anybody could sign up), it randomly sampled 15% of its customers and told then that if they enrolled in the loyalty program when it opened, they would be entered into a drawing to receive one of 10 Amazon gift cards worth €100 each.

To obtain the data from this experiment, run the following code:

```{r}
loyalty <- get_ECI_data('module-7-tutorial')
loyalty
```

The variables are:

* `customer`: a unique identifier for each customer
* `Z`: whether the customer received the incentive to join the loyalty program
* `D`: whether the customer joined the loyalty program
* `Y`: revenue within the first 3 months after the launch of the loyalty program (for the sake of simplicity, assume that customers could only enroll when the program was first launched)

The company wants to know a few things:

* Was revenue higher among customers who received the incentive? (This is the intent-to-treat effect of the incentive on revenue, or ITT)
* Were customers who were offered the incentive to join, in fact, more likely to join the loyalty program? (This is the intent-to-treat effect of the incentive on joining the loyalty program, or ITTd)
* How much did joining the loyalty program increase revenue among the subset of customers who joined the loyalty program because of the incentive, but wouldn't have joined the loyalty program if not offered the incentive? (This is the complier average causal effect, CACE, which is also called the local average treatment effect, LATE)
* Was the incentive cost effective?

The company believes or is willing to assume the following things:

* Customers who stand to benefit the most from joining the program are more likely to join it. That is, the change in revenue from joining the loyalty program is probably correlated with the decision to join the loyalty program.
* There are "always-takers"—some customers who were offered the incentive to join and subsequently joined the loyalty program would have joined even without the incentive.
* There are "never-takers"—some customers who were  not offered the incentive and did not join, still would not have joined, even with the incentive.
* There are "compliers"—some customers who received the incentive and joined the loyalty program would not have joined if they hadn't received the incentive.
* There are no "refusers"—none of the customers who received the incentive became less likely to join as a result of receiving the incentive. (This is the monotonicity assumption in two-sided non-compliance problems.)
* The usual assumptions about there being no spillovers between customers (SUTVA), etc.

## Structure of the problem

The following DAG shows why estimation of the causal effect of joining the loyalty program on revenue cannot be estimated through standard methods.

```{r M8-loyalty-DAG, echo = FALSE}
plot((dagitty::dagitty('dag {
D [exposure,pos="-0.500,0.000"]
Y [outcome,pos="1.500,0.000"]
Z [pos="-1.500,1.000"]
u [latent,pos="0.500,-1.000"]
D -> Y
Z -> D
u -> D
u -> Y
}')))
```

The unobserved variable `u` affects both revenue (`Y`) and enrollment in the loyalty program (`D`). However, because the company is willing to assume that the incentive (`Z`) affects enrollment, but has no direct effect on revenue (`Y`), we can use the offer of the incentive as an "instrumental variable" for enrollment (`D`). 

We will estimate the effects of interest using three different methods, in order to illustrate that these all yield the same estimated CACE/LATE.

## Estimation

First, we will estimate the intent-to-treat effects of the incentive on 1) revenue, and 2) enrollment. Let's start with ITT, the effect of the incentive on revenue:

```{r}
loyalty_ITT <-
	loyalty %>%
	group_by(Z) %>%
	summarise(Y = mean(Y)) %>%
	pivot_wider(names_from = Z, values_from = Y, names_prefix = 'Z_')
loyalty_ITT
```

`Z_0` is the average revenue (`Y`) among customers who were not offered the incentive to join the loyalty program, and `Z_1` is the average revenue among customers who were offered the incentive. 

The difference between the two values is our estimate of the intent-to-treat effect (ITT) of the incentive on revenue:

```{r}
ITT <- 
	loyalty_ITT %>%
	mutate(ITT = Z_1 - Z_0) %>%
	getElement('ITT')
ITT
```

Revenue in the first 3 months was €`r signif(ITT, 2)` higher among customers who were offered the incentive.

Next, we will calculate ITTd, the intent-to-treat effect of the incentive on the decision to join the program.

```{r}
loyalty_ITTd <-
	loyalty %>%
	group_by(Z) %>%
	summarise(D = mean(D)) %>%
	pivot_wider(names_from = Z, values_from = D, names_prefix = 'Z_')
loyalty_ITTd
```

Both values indicate the share of customers who joined the loyalty program; `Z_0` is the average among customers who were not offered the incentive, and `Z_` is the average among customers who were ofference the incentive. 

Their difference is an estimate of the intent-to-treat effect of the incentive on "compliance" (i.e., joining the program); the ITTd: 

```{r}
ITTd <-
	loyalty_ITTd %>%
	mutate(ITTd = Z_1 - Z_0) %>%
	getElement('ITTd')
ITTd
```

The share of customers joining the program was `r signif(ITTd*100, 2)`% higher among customers who were offered the incentive to join. 

By taking the ratio of these two estimates, `ITT / ITTd`, we can obtain an estimate of the CACE/LATE:

```{r}
ITT / ITTd
```

Among customers who joined the loyalty program because they were offered the incentive (so they would not join the program without the incentive), revenue increased by about €`r signif(ITT/ITTd, 2)` due to joining the loyalty program.

## Was the incentive worth it?

In exchange for €1000 (the cost of ten gift cards) the company was able to increase revenue among a subset of customers. There are a few ways to answer the question of whether the incentive was profitable. All involve 1) calculating the total increase in revenue due to joining the loyalty program because of the gift card incentive, and 2) comparing that amount to the €1000 cost of the promotion.

So what was the revenue lift? To estimate this, we first need to estimate the number of customers who joined the program because of the incentive. That is, we need to estimate the proportion of compliers among the company's customers.

### How many compliers are there?

The ITTd can serve as an estimate of the share of compliers (think of it as estimating the share of people who switched from `D == 0` to `D == 1` due to the incentive). Alternatively, and for the purposes of this tutorial, we can estimate the share of never-takers, always-takers, and compliers directly. 

To do this, we first estimate the proportion of customers who did not join the program, among the subset who were not offered the incentive. This group includes both never-takers and compliers (stop and think through why that is the case), so we'll call this quantity `nt_and_c`. 

```{r}
nt_and_c <- sum(loyalty$Z == 0 & loyalty$D == 0) / sum(loyalty$Z == 0)
nt_and_c
```

Second, we can estimate the proportion of customers who joined the program, among the subset who were not offered the incentive. This group includes both always-takers and defiers---but since we assume there are no defiers (*monotonicity*), this quantity is simply the share of always-takers among the company's customers. The share of always-takers is equal to `1 - nt_and_c`, but we can also calculate it directly:

```{r}
at <- sum(loyalty$Z == 0 & loyalty$D == 1) / sum(loyalty$Z == 0)
at
```

Third, we can estimate the proportion of customers who joined the program, among those who were offered the incentive. This group includes a mix of always-takers and compliers. (Again, stop and think through why this is).

```{r}
at_and_c <- sum(loyalty$Z == 1 & loyalty$D == 1) / sum(loyalty$Z == 1)
at_and_c
```

Fourth, we can estimate the proportion of customers who did not join the program, among the subset who were offered the incentive. Again, because we assume there are no defiers, this group contains only never-takers:

```{r}
nt <- sum(loyalty$Z == 1 & loyalty$D == 0) / sum(loyalty$Z == 1)
nt
```

We can then derive the share of compliers using either of the two methods below:

```{r}
at_and_c - at
nt_and_c - nt
```

This estimate is also the ITTd, as mentioned above.

```{r}
ITTd
```

Compliers therefore make up about `r round(100*(at_and_c - at), 0)`% of the company's customers. Among the `r sum(loyalty$Z == 1)` who were offered the incentive, `r round(ITTd * sum(loyalty$Z == 1))` joined because of the incentive. 

```{r}
compliers_offered_incentive <- ITTd * sum(loyalty$Z == 1)
compliers_offered_incentive
```

### Was the incentive profitable?

Back to our original question of whether the increase in revenue from the incentive was worth its cost. Recall that the average change in revenue due to joining the loyalty program is given by our estimate of the CACE, which we estimated as the ratio of `ITT/ITTd`:

```{r}
ITT/ITTd
```

Multiplying this average by the number of compliers who received the incentive yields an estimate of the total increase in revenue due to the incentive:

```{r}
compliers_offered_incentive * ITT / ITTd
```

The company spent €1000 on the incentive, but in return, revenue increased by €`r round(compliers_offered_incentive * ITT / ITTd,0)`. Ignoring all other costs, the incentive seems to have been profitable.

A final point before moving on to regression. Another way to estimate the revenue lift attributed to the promotion is to multiply the ITT (the average revenue lift among customers offered the incentive) by the number of customers who were offered the incentive.

```{r}
sum(loyalty$Z == 1) * ITT
```

This gives the same answer as the number of compliers offered the incentive times the CACE/LATE. 

```{r}
compliers_offered_incentive * ITT / ITTd
sum(loyalty$Z == 1) * ITT
```

The two are the same, and this illustrates that ITT effects are driven entirely by the behavior of compliers. This is a somewhat subtle point, but it's an important one to remember when running field experiments around compliance promotions. Any ITT or CACE/LATE effects you measure arise due to the behavior of the subset of customers who, without the incentive, would not comply with the promotion, but with the incentive, do comply. That means the treatment effect is defined in the context of the incentive that induces compliance (i.e., the instrumental variable). In our example, the treatment effect we just estimated is valid only for the incentive that was offered. Had we offered a chance to win one of 20 €50 Amazon gift cards, the treatment effect might have been different (even though the total cost of the promotion is the same for us).


## Estimating CACE/LATE via IV regression

When our excluded variable is limited to just two values (i.e., when the instrument is binary), then we can also estimate the CACE/LATE using instrumental variables (IV) regression (Note: if the instrument is not binary valued, then IV regression yields a weighted average of CACEs that lacks clear interpretation). First, we will carry out two-stage-least-squares (2SLS) regression manually, then we will use the `ivreg` function from the AER package to perform the regression with one call.

The first step in 2SLS is to obtain predicted values of `D` for each customer. We can obtain these by regressing `D` on `Z` (using OLS).

```{r}
reg_DZ <- lm(D ~ Z, data = loyalty)
summary(reg_DZ)
```

(Notice that the coefficient for `Z` is equal to the ITTd.) To obtain a vector of predicted values of `D`, which we will call `D_hat`, we use the `predict` function:

```{r}
D_hat <- predict(reg_DZ)
```

Next, we will regress `Y` on `D`, but rather than using the real values of `D`, we will instead use the values of `D_hat` we just estimated.

```{r}
loyalty_with_D_hat <-
	loyalty %>%
	mutate(D_hat = D_hat)
loyalty_with_D_hat
reg_Y_D_hat <- lm(Y ~ D_hat, data = loyalty_with_D_hat)
summary(reg_Y_D_hat)
```

The coefficient for `D_hat` is equal to our earlier estimate of the CACE/LATE obtained via the ratio of `ITT/ITTd`:

```{r}
ITT / ITTd
```

In practice, we would normally use the `ivreg` function from the AER package to perform IV regression. 

```{r}
library(AER)
```

To specify the regression, we will use the `|` symbol when specifying the formula:


```{r}
reg_Y_D_iv <- ivreg(Y ~ D | Z, data = loyalty)
```

The formula `Y ~ D | Z` means "regress `Y` on `D` after replacing `D` with estimates from a regression of `D` on `Z`." Or maybe more simply, "use `Z` as an instrument for `D` in a regression of `Y` on `D`."

Finally, we can see that this also yields the same estimate for the CACE/LATE:

```{r}
summary(reg_Y_D_iv)
ITT / ITTd
```



