---
title: "PS2: Satisfaction data"
output: 
  rmarkdown::html_document:
    fig_width: 6
    fig_height: 4
    mathjax: null
    theme: cerulean
    highlight: pygments
    self_contained: false
vignette: >
  %\VignetteIndexEntry{P2-satisfaction}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

<style>
pre > code.sourceCode a {
  display: none;
}
pre {
  border-width: 0px;
}
code {
  border-width: 0px;
}
</style>


```{r, include = FALSE}
if (require(canvasapicore)) {
  knitr::opts_knit$set(upload.fun = function(file) {
    canvasapicore::load_token_and_domain()
    resp <- cnvs::cnvs_upload("/api/v1/folders/601289/files", file)
    structure(resp$url, XML = resp)
  })
}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Load packages

```{r setup, message = FALSE, warning = FALSE}
library(ECICourse)
library(tidyverse)
```


Get data for this problem

```{r}
satisfaction <- get_ECI_data('problem-set-2-satisfaction')
```

The data frame `satisfaction` describes the results of an imagined field experiment  (that I simulated in R just  for this problem set). The  column `D`  indicates  assignment to the treatment  condition (`D == 1`) or control condition (`D == 0`). `Y` contains the observed customer  satisfaction rating on a scale of 1--5. Each row represents the assignment and rating of a different customer.

```{r}
satisfaction %>%
  select(D, Y)
```

The data frame `satisfaction` also contains two columns that you would never observe in the real world. For each customer, the column `Y1` contains the potential outcome under treatment and  `Y0`  the potential outcome under control.

```{r}
satisfaction
```

Again, you would never observe all of `Y1` and `Y0` in the real world. Instead, you only observe `Y == D * Y1 + (1 - D) * Y0`. But I have included  these columns so that, for the purposes of this problem set, you can calculate theoretical quantities that we believe exist, but never observe directly. 

For example, the true average treatment effect (ATE) can be calculated by taking the average difference between `Y1` and `Y0`:

```{r}
satisfaction %>%
  summarise(true_ATE = mean(Y1 - Y0))
```

Use the following code as a starting point to answer the questions in this problem set related to this vignette.

```{r results = 'hide', message = FALSE}
# average values of (observed) Y among customers with either D == 0 or D == 1
satisfaction %>%
  group_by(D) %>%
  summarise(Y = mean(Y)) %>%
  pivot_wider(names_prefix = 'avg.Y|D=', names_from = 'D', values_from = 'Y') %>%
  mutate(naive.ATE = `avg.Y|D=1` - `avg.Y|D=0`)

# average values of Y1, Y0, delta = Y1 - Y0, and Y, conditional on D
satisfaction %>%
  group_by(D) %>%
  summarise(avg.Y = mean(Y),
            avg.Y1 = mean(Y1),
            avg.Y0 = mean(Y0),
            avg.delta = mean(Y1 - Y0))
```


